Model Search in Relational Domains

### 2016-10-22 deductions.tex
sync

\section{Abstract Model Search}
%
\Omit {
\begin{figure}[t]
\scriptsize
\begin{tabular}{l|l|l}
\hline
C program & SSA & Octagon \\
\hline
\begin{lstlisting}[mathescape=true,language=C]
int main() {
 unsigned x, y;
 __CPROVER_assume(x==y);
 x++;
 assert(x==y+1);
}
\end{lstlisting}
&
\begin{minipage}{4.40cm}
$\begin{array}{l@{\,\,}c@{\,\,}l}
SSA &\iff& ((g0 == TRUE) \land \\
    &    & (cond == (x == y)) \land \\
    &    & (g1 == (cond \&\& guard0)) \land \\
    &    & (x' == 1u + x) \land \\
    &    & (x' == 1u + y || !1))
\end{array}$
\end{minipage}
&
\begin{minipage}{3.75cm}
$\begin{array}{l@{\,\,}c@{\,\,}l}
C &\iff& ((x' > 1) \land (-x'-y < -2) \land \\
  &    & (-x-x' < -2) \land (y-x' < 0) \land \\                                                                
  &    & (x-x' < 0) \land (y > 0) \land \\
  &    & (x > 0) \land (-x'-y < 0) \land \\
  &    & (x+y > 1) \land (y-x < 1) \land \\
  &    & (x'-y < 2) \land (x-y < 1) \land \\
  &    & (x+y > 0) \land (x+x' > 0) \land \\
  &    & (x'-x < 2))
\end{array}$
\end{minipage}
\\
\hline
\end{tabular}
\caption{C Program, corresponding SSA and Octagonal Inequalities}
\label{ssa}
\end{figure}
}
%
%    
\begin{algorithm2e}[t]
\DontPrintSemicolon
\SetKw{return}{return}
\SetKwData{sat}{sat}
\SetKwData{conflict}{conflict}
\SetKwData{unsat}{unsat}
\SetKwData{unknown}{unknown}
\SetKwData{true}{true}
\SetKwInOut{Input}{input}
\SetKwInOut{Output}{output}
\SetKwFor{Loop}{Loop}{}{}
\SetKw{KwNot}{not}
\begin{small}
\Input{A program in the form of a set of abstract transformers $\abstransset$,
a propagation trail $\trail$, and a reason trail $\reasons$.}
\Output{\sat or \conflict or \unknown}
$\worklist \leftarrow \initworklist_{\propheur}(\abstransset)$ \;
\While{$!\mathit{worklist.empty}()$} 
{
  $\abstransel{\subdomain} \leftarrow \mathit{worklist.pop}()$ \; 
  $\newdeductions \leftarrow \onlynew(\abstransel{\subdomain}(\abs(\trail)))$\;
  \uIf{$\newdeductions = \bot$} {
    $\reasons[\bot] \leftarrow \abstransel{\subdomain}$ \;
    $\mathit{worklist.clear}()$ \;
    \return \conflict \;
  }
  \uElse
  {
%    $\newdeductions=\decomp(\absval)\setminus\decomp(\abs(\trail))$\;
    $\trail \leftarrow \trail \cdot \decomp(\newdeductions)$ \; 
    $\reasons[|\trail|] \leftarrow \abstransel{\subdomain}$ \;
    $\updateworklist_{\propheur}(\newdeductions, \worklist)$ \; 
  }
}
\lIf{$\abs(\trail)$ is $\gamma$-complete} {
  \return \sat;
}
 \return \unknown \;

@@

\paragraph {\textbf{Lazy closure computation}} 
Computing closure for relational domains, such as octagons is extremely
expensive~\cite{pldi15}.  An advantage of our formalism in 
Eq.~(\ref{eq:at2}) is that we can compute the
\emph{closure} operation for relational domains in a lazy manner. To
this end, we construct a $\subdomain=\makesubdomain_\domain(\subvars)$
which is sufficient to perform one step of the closure operation each
time the abstract transformer is applied.
%
For example, let us consider $\domain=\octagons[\{x,y,z\}]$ and
$\subvars=\{y\}$. An octagonal inequality relates at 
most two variables. Thus it is sufficient to consider the subdomain
$\makesubdomain_\domain(\{y\})=\octagons[\{y\}]\cup\octagons[\{x,y\}]\cup\octagons[\{y,z\}]$,
which will compute the one-step transitive relations of~$y$ with each
of the other variables. 
%
Hence, when the abstract deduction transformer is applied we do not
compute the full closure in the full domain,
but we compute only a single step of the closure in a restricted
domain, which makes single deduction steps more efficient.
%
If $\abstransel{\subdomain}$ deduces new information about $x$ or $z$
then the next step of the closure will be computed by the worklist
mechanism of Algorithm~\ref{Alg:ms} that we describe next.

@@

Algorithm~\ref{Alg:ms} presents the deduction phase $\deduce$ in 
abstract model search procedure.  The input to the algorithm is 
the set of abstract transformers, a propagation trail and a reason 
trail.  Additionally, the procedure $\deduce$ is parameterised with 
a propagation heuristic. 
%
The algorithm maintains a  {\em worklist}, which is a queue that contains 
abstract deduction transformers.  
The propagation heuristics provides two 
functions $\initworklist$ and $\updateworklist$.
The order of the elements in the worklist and the subdomains $\subdomain$
attached to each of the abstract deduction transformers
determine the propagation strategy (forward, backward, multi-way).
%
The {\em forward} propagation strategy initialises the worklist with
abstract deduction transformers that contains constants in the
right-hand side and the subdomain is constructed via $\makesubdomain$
from variables in left-hand side of the equality constraints
originating from the assignment statements in the program.
%
The {\em backward} iteration strategy intialises the worklist with the assertions and the subdomain is constructed from the right-hand side variables.
%
The {\em multi-way} iteration strategy initialises the worklist 
with set of all transformers and each subdomain contains the variables occurring in the respective transformers.

Depending on the propagation heuristics, $\updateworklist$ adds
abstract deduction transformers to the worklist that contain variables
that appear in $\newdeductions$, and it updates the subdomains of
the abstract deduction transformers in the worklist 
to include the variables in $\newdeductions$.

The function
$\onlynew(\absval)=\bigsqcap(\decomp(\absval)\setminus\decomp(\abs(\trail)))$
is used to filter out all meet irreducibles that are already on the trail
in order to obtain only new deductions when applying the abstract deduction transformer.

If $\bot$ is deduced we return \textsf{conflict}.
Otherwise, when eventually a fixed point has been reached, i.e.\ the worklist is empty, then the abstract value $\abs(\trail)$ is checked whether it is 
$\gamma$-complete~\cite{dhk2013-popl}. It is $\gamma$-complete if all concrete values in $\gamma(\abs(\trail))$ satisfy $\formula$.
Otherwise, the algorithm returns \textsf{unknown} and the 
model search procedure makes a new decision.

>>>>

\section{Abstract Model Search}
%
\Omit {
\begin{figure}[t]
\scriptsize
\begin{tabular}{l|l|l}
\hline
C program & SSA & Octagon \\
\hline
\begin{lstlisting}[mathescape=true,language=C]
int main() {
 unsigned x, y;
 __CPROVER_assume(x==y);
 x++;
 assert(x==y+1);
}
\end{lstlisting}
&
\begin{minipage}{4.40cm}
$\begin{array}{l@{\,\,}c@{\,\,}l}
SSA &\iff& ((g0 == TRUE) \land \\
    &    & (cond == (x == y)) \land \\
    &    & (g1 == (cond \&\& guard0)) \land \\
    &    & (x' == 1u + x) \land \\
    &    & (x' == 1u + y || !1))
\end{array}$
\end{minipage}
&
\begin{minipage}{3.75cm}
$\begin{array}{l@{\,\,}c@{\,\,}l}
C &\iff& ((x' > 1) \land (-x'-y < -2) \land \\
  &    & (-x-x' < -2) \land (y-x' < 0) \land \\                                                                
  &    & (x-x' < 0) \land (y > 0) \land \\
  &    & (x > 0) \land (-x'-y < 0) \land \\
  &    & (x+y > 1) \land (y-x < 1) \land \\
  &    & (x'-y < 2) \land (x-y < 1) \land \\
  &    & (x+y > 0) \land (x+x' > 0) \land \\
  &    & (x'-x < 2))
\end{array}$
\end{minipage}
\\
\hline
\end{tabular}
\caption{C Program, corresponding SSA and Octagonal Inequalities}
\label{ssa}
\end{figure}
}
%
%    
\begin{algorithm2e}[t]
\DontPrintSemicolon
\SetKw{return}{return}
\SetKwData{sat}{sat}
\SetKwData{conflict}{conflict}
\SetKwData{unsat}{unsat}
\SetKwData{unknown}{unknown}
\SetKwData{true}{true}
\SetKwInOut{Input}{input}
\SetKwInOut{Output}{output}
\SetKwFor{Loop}{Loop}{}{}
\SetKw{KwNot}{not}
\begin{small}
\Input{A program in the form of a set of abstract transformers $\abstransset$,
a propagation trail $\trail$, and a reason trail $\reasons$.}
\Output{\sat or \conflict or \unknown}
$\worklist \leftarrow \initworklist_{\propheur}(\abstransset)$ \;
\While{$!\mathit{worklist.empty}()$} 
{
  $\abstransel{\subdomain} \leftarrow \mathit{worklist.pop}()$ \; 
  $\newdeductions \leftarrow \onlynew(\abstransel{\subdomain}(\abs(\trail)))$\;
  \rmcmt{can we avoid onlynew or atleast change the name?}
  \uIf{$\newdeductions = \bot$} {
    $\reasons[\bot] \leftarrow \abstransel{\subdomain}$ \;
    $\mathit{worklist.clear}()$ \;
    \return \conflict \;
  }
  \uElse
  {
%    $\newdeductions=\decomp(\absval)\setminus\decomp(\abs(\trail))$\;
    $\trail \leftarrow \trail \cdot \decomp(\newdeductions)$ \; 
    $\reasons[|\trail|] \leftarrow \abstransel{\subdomain}$ \;
    $\updateworklist_{\propheur}(\newdeductions, \worklist)$ \; 
  }
}
\lIf{$\abs(\trail)$ is $\gamma$-complete} {
  \return \sat;
}
 \return \unknown \;

@@

\paragraph {\textbf{Lazy closure computation}} 
Computing closure for relational domains, such as octagons is extremely
expensive~\cite{pldi15}.  An advantage of our formalism in 
Eq.~(\ref{eq:at2}) is that
%we can compute the
the \emph{closure} operation for relational domains can be computed 
in a lazy manner. To this end, we construct a subdomain 
$\subdomain=\makesubdomain_\domain(\subvars)$ which is sufficient 
to perform one step of the closure operation each
time the abstract \rmcmt{deduction} transformer is applied.
%
For example, let us consider $\domain=\octagons[\{x,y,z\}]$ and
$\subvars=\{y\}$. An octagonal inequality relates at 
most two variables. Thus it is sufficient to consider the subdomain
$\makesubdomain_\domain(\{y\})=\octagons[\{y\}]\cup\octagons[\{x,y\}]\cup\octagons[\{y,z\}]$,
which will compute the one-step transitive relations of~$y$ with each
of the other variables. 
%
\rmcmt{Only if a abstract deduction transformer subsequently makes new deductions 
on $x$ or $z$, then the next step of the closure will be computed through 
the subdomain $\octagons[\{x,z\}]$.}
\Omit{
Hence, when the abstract deduction transformer is applied we do not
compute the full closure in the full domain,
but we compute only a single step of the closure in a restricted
domain, which makes single deduction steps more efficient.
}
\rmcmt{Hence, an application of abstract deduction transformer does not 
compute the full closure in the full domain, but compute only a 
single step of the closure in a restricted domain, which makes each 
deduction step more efficient.  Thus, we delay the closure operation 
until the point where it is absolutely necessary to compute it, thereby 
increasing the efficiency of the deductions.}
%
\Omit{
If $\abstransel{\subdomain}$ deduces new information about $x$ or $z$
then the next step of the closure will be computed by the worklist
mechanism of Algorithm~\ref{Alg:ms} that we describe next.
}
%

@@

Algorithm~\ref{Alg:ms} presents the deduction phase $\deduce$ in 
abstract model search procedure.  The input to the algorithm is 
the set of abstract transformers, a propagation trail and a reason 
trail.  Additionally, the procedure $\deduce$ is parameterised with 
a propagation heuristic ($\propheur$). 
%
The algorithm maintains a  {\em worklist}, which is a queue that contains 
abstract deduction transformers.  The propagation heuristics provides two 
functions $\initworklist$ and $\updateworklist$.
The order of the elements in the worklist and the subdomains $\subdomain$
attached \rmcmt{or associated} to each abstract deduction transformers 
\rmcmt{($\abstransel{\subdomain}$)} determine the propagation strategy 
(forward, backward, multi-way).
%
The {\em forward} propagation strategy initialises the worklist with
abstract deduction transformers that contains constants in the
right-hand side and the subdomain is constructed via $\makesubdomain$
from variables in left-hand side of the equality constraints
originating from the assignment statements in the program.
%
The {\em backward} propagation strategy intialises the worklist 
with the assertions and the subdomain is constructed from the 
right-hand side variables.
%
The {\em multi-way} propagation strategy initialises the worklist 
with set of all transformers and each \rmcmt{each or the corresponding ?} 
subdomains contains the variables occurring in the respective transformers.
%
Depending on the propagation heuristics, $\updateworklist$ adds
abstract deduction transformers to the worklist that contain variables
that appear in $\newdeductions$, and it updates the subdomains of
the abstract deduction transformers in the worklist 
to include the variables in $\newdeductions$.
%
The function
$\onlynew(\absval)=\bigsqcap(\decomp(\absval)\setminus\decomp(\abs(\trail)))$
is used to filter out all meet irreducibles that are already on the trail
in order to obtain only new deductions when applying the abstract deduction transformer.
%
\Omit{
If $\bot$ is deduced we return \textsf{conflict}.
Otherwise, when eventually a fixed point has been reached, i.e.\ the worklist is empty, then the abstract value $\abs(\trail)$ is checked whether it is 
$\gamma$-complete~\cite{dhk2013-popl}. It is $\gamma$-complete if all concrete values in $\gamma(\abs(\trail))$ satisfy $\formula$.
Otherwise, the algorithm returns \textsf{unknown} and the 
model search procedure makes a new decision.    
}
\rmcmt{If $\abstransel{\subdomain}$ deduce $\bot$, then 
the procedure $\mathit{deduce}$ returns \textsf{conflict}.
Otherwise, when a fixed point is reached, i.e.\ 
the worklist is empty, then the abstract value $\abs(\trail)$ is checked 
for $\gamma$-completeness~\cite{dhk2013-popl}.  If it is indeed 
$\gamma$-complete, then $\mathit{deduce}$ returns \sat.  Otherwise, the 
algorithm returns \textsf{unknown} and the abstract model search 
procedure makes a new decision.    
}

### 2016-10-22 introduction.tex

Abstract Conflict Driven Clause Learning} (ACDCL)~\cite{dhk2013-popl} is one
such generalisation of CDCL.  The partial satisfying assignments in CDCL
algorithm are generalized to be chosen from a given abstract domain.  The
unit rule is the best abstract transformer, Boolean Constraint Propagation
(BCP) computes a greatest fixed-point over the abstract domain, decision is
dual widening and clause learning can be viewed as synthesising an abstract
transformer for negation.  This enables an ACDCL-based analyser to refine an
imprecise analysis and thus prevents enumeration behaviour.  ACDCL has so
far been instantiated as a decision procedure for floating-point
arithmetic~\cite{DBLP:journals/fmsd/BrainDGHK14} and a program analyser that
reasons using interval~\cite{tacas12}.  Hence, the existing ACDCL instances
are limited to bounds checking in numerical programs.

>>>>

\emph{Abstract Conflict Driven Clause Learning} (ACDCL)~\cite{dhk2013-popl}
is one such generalisation of CDCL.  The partial satisfying assignments in
CDCL algorithm are generalized to be chosen from a given abstract domain. 
The unit rule is the best abstract transformer, Boolean Constraint
Propagation (BCP) computes a greatest fixed-point over the abstract domain,
decision is dual widening and clause learning can be viewed as synthesising
an abstract transformer for negation.  This enables an ACDCL-based analyser
to refine an imprecise analysis and thus prevents enumeration behaviour. 
ACDCL has so far been instantiated as a decision procedure for
floating-point arithmetic~\cite{DBLP:journals/fmsd/BrainDGHK14} and a
program analyser that reasons using intervals~\cite{tacas12}.  Hence, the
existing ACDCL instances are limited to bounds checking in numerical
programs.
